DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-12-22 2015CompleteDerivation/completeDerivation.tex
A global pass to make the language more homogeneous throughout

We can also conceptually combine two different processes into a single one. That is, having a way to measure quantity $x$ and a way to measure quantity $y$ gives us a way to measure the combination $(x,y)$. For ensembles, if we can measure the marginal distribution $\rho_x(x)$ and the marginal distribution $\rho_y(y)$, we know the joint distribution $\rho(x,y)$ has to be compatible with both. Note, though, that this does not provide a way to fully measure $\rho(x,y)$ as we know nothing about the correlation between the two variables.\footnote{The quantum case is similar: we measure marginal distributions and rule out states that are incompatible with them.} Formally, the states compatible with the outcomes of the combined process will be the intersections of the states compatible with each pair of outcomes of the original processes.

We can also coarsen a single process. That is, having a way to measure $(x,y)$ gives us a way to measure $x$ alone (or any $f(x,y)$). Formally, the outcomes of the coarsened process are given by performing the union of some outcomes of the original process.

@@

With our state space defined, deterministic and reversible evolution corresponds to a bijective map between initial and final states. Moreover, a physical process that can distinguish final states is also a physical process that can distinguish initial states. That is: the topology is mapped and preserved by the deterministic and reversible evolution (i.e. initial and final states are equally distinguishable). The evolution is then a self-homeomorphism on the state space.

@@

The generality of this result explains why in physics one always assumes functions to be well behaved. As the result was derived from our notion of physical distinguishability, this is not a matter of practical convenience. Suppose we were able to prepare a force field that was zero everywhere in space except at a single point. This gives us a way to tag a specific point. But it also allows us to create an outcome compatible with only a single point. A finite precision measurement of the force provides us infinite precision of space, which we ruled out.\footnote{This assumes that we are able to position the probe perfectly, which we could do if we were able to manipulate forces at that precision.} This is what the math is telling us, that if we said before that outcomes can't distinguish isolated points (i.e. the topology is second countable) then neither can maps. It is physical consistency that limits us to continuous functions.

@@

To identify and name states one uses a set of quantities, typically numbers. For example, the orbital of an electron in a hydrogen atom is identified by the quantum numbers $n$, $l$, $m$ and $s$. We call each of these quantities \emph{state variables}. We call a \emph{possibility} a possible value that can be taken by a state variable. We call a state variable \emph{discrete} or \emph{continuous} if the possibilities are integer or real numbers respectively. From now on, we are going to consider state spaces whose states can be identified, at least within a region, by a finite set of discrete and continuous state variables (i.e. the state space is locally isomorphic to $\bigcup\limits_{1 \leq i \leq n} \mathbbm{R}^{m_i}$).\footnote{TODO: Is there a more physically meaningful requirement? Could we have a set of states that are not locally described by a set of quantities?}

We purposely use the term state variables instead of coordinates (even though that's what they are mathematically) as it would create confusion with space-time coordinates. We also avoid the term observable or measurable, as not all state variables may be directly physically tangible (e.g. conjugate momentum in a gauge theory). The only requirements for state variables is that they identify states. This means the possible values for state variables are physically distinguishable and are defined at equal time (since states are defined and mapped at a particular time). \footnote{This is the main reason that a quantity like velocity is not a suitable state variable, as it is  defined over an interval of time. Therefore velocity is always physically well defined but is not a state variable in general, while conjugate momentum is always a state variable but is not physically well defined by itself (it requires the vector potential to be specified as well). As we'll see later, if there exists a one to one map between velocity and conjugate momentum they can both be physically well defined state variables.}

@@

When combining multiple state variables, it is important to understand how they relate to each other. Consider the orbital of an electron in a hydrogen atom, which is identified by the quantum numbers $n$, $l$, $m$ and $s$. For each combination of $n$, $l$ and $m$, the spin $s$ can have two values. The choice of spin is independent from the rest. The choice of $l$, though, depends on the choice of $n$: for $n=1$ only $l=0$ is available; for $n=2$ we can choose $l=0$ or $l=1$. The choices are not independent. That is: two or more variables are independent if there always exists a state for any possible combination, if the total number of states is the cartesian product of the possibilities of each variable. We call \emph{state vector} a collection of independent state variables that fully identify a state.

@@

The second approach is to transport the state variables. The idea is to keep the connection to the initial state by labeling the future state by the original value of $q^i$, instead of by the future values. For example, the evolved state variable will not tell us the current value of pressure, but the one for the initial conditions. So we introduce a new set of variables for which $\hat{q}^i(\hat{\mathcal{s}})=q^i(\mathcal{s})$, which we can always do as the evolution is deterministic and reversible. The level sets of $\hat{q}^i$ are the evolved level sets of $q^i$ (e.g. all the states that started with a particular value for pressure) which allows us to study how groups of states evolve in time. Given that the value of the transported state variables does not change during evolution, and that it is unique for each initial state, transported state variables also provide a way to label the trajectories themselves.

Evolved state variables are useful to write equations of motions, study how physical quantities change, form a physical picture of what happens. Transported state variables are useful to write invariants, study state space trajectories, form a geometric picture for the state space.\footnote{One should not confuse evolved/transported state variable with active/passive transformations or with Schroedinger/Heisenberg pictures. In those cases, the choice is between changing the state or the coordinates/observables. In our case, the state is always changing. The choice is between tracking the change using different state variables or different values.}

@@

To show the usefulness of transported state variables, we can prove the following:

@@

The issue is that while deterministic evolution will map states to states, not all bijective state to state maps can be considered deterministic and reversible. This is a crucial point where this work is different from more usual treatments of dynamical systems. If one focuses only on the states themselves and counts them, it reaches the conclusion that a range of $1$m contains the same number of initial conditions of $1$Km: they have the same infinite number of elements. This conclusion has a number of problems: it does not match our physical intuition; it's not consistent with all types of topology; it does not explain why the laws of evolution are very often, not only continuous, but differentiable. In short: a bijection does not capture the entirety of the physical description.

If we work with a discrete topology then, by all means, the way to measure the size of a set is to count the elements in a set. But if we use the standard topology on $\mathbbm{R}$, isolated points have measure zero: what defines the number of initial conditions is the range of the state variables. As we'll see, the geometrical structures in Hamiltonian/Lagrangian classical/quantum mechanics are there to allow us to properly count states and, more specifically, possibilities along independent state variables. If we think that counting states is equivalent to counting elements in the set, we miss the whole point of that geometrical structure.

@@

Given that scientific reductionism (i.e. the idea of reducing physical systems and interactions to the sum of their constituent parts in order to make them easier to study) is at the heart of fundamental physics, we now explore how to characterize a system in terms of its components. That is, we want to study the relationship between the state space of a composite system and the state space of its parts. In general, this is quite a complicated thing to do, which requires intimate knowledge of the system at hand. So we simplify our problem and study a material made of infinitesimal homogeneous parts. What we'll find is that under the additional assumption that the material is infinitesimally reducible (i.e. its state is equivalent to the states of its parts) and that each part undergoes deterministic and reversible evolution, the motion is suitably described by the standard framework of classical Hamiltonian mechanics.

\subsection{Homogeneous decomposable systems and vector spaces}
The notion that a system is decomposable means the states are equipped with a rule of composition that allows one to write $\mathcal{c}=\mathcal{c}_1+\mathcal{c}_2$: the composite system is the sum of its parts. For example, the state of a ball is equal to the state of its top and bottom parts.

@@

\begin{justification}
We claim $\mathcal{C}$ is an additive monoid. There exists a law of composition $+ : \mathcal{C} \times \mathcal{C} \rightarrow \mathcal{C}$ that takes two states and returns one that is the physical composition of the two. The domain and codomain match because the material is homogeneous. The law is commutative $\mathcal{c}_1 +\mathcal{c}_2 = \mathcal{c}_2+\mathcal{c}_1$ and associative $(\mathcal{c}_1 + \mathcal{c}_2) + \mathcal{c}_3 = \mathcal{c}_1 + (\mathcal{c}_2 + \mathcal{c}_3)$, as it does not matter in what order we physically compose the parts. There exists a unique zero element $\mathcal{c} + 0 = \mathcal{c}$ and it represents the physically empty state (i.e. no amount of material). $\mathcal{C}$ is an additive monoid by definition.

@@

We will call a \emph{particle} the smallest amount of a material to which we can assign an independent state. For example, a photon is a particle of light (it is the smallest amount of light we can describe), an infinitesimal amount of water is treated classically as a point particle (the smallest amount for a continuous fluid).\footnote{We use this definition as it allows us to retroactively talk on somewhat equal grounds about classical point-particles and quantum particles, and underscore how physically (and mathematically) they are very similar yet very different objects.}

We will call a \emph{classical material} one that is homogeneous and infinitesimally reducible. That is, we can keep decomposing the system indefinitely into smaller and smaller parts, each with a well defined state. In this case, the particles are the infinitesimal parts given by the limit of this process of recursive decomposition. Giving the state of the whole system is equivalent to giving the states of all particles. Given the state space $\mathcal{S}$ for the infinitesimal parts, which we assume consistently with \ref{continuous_state_space} to be a manifold, a composite state $\mathcal{c} \in \mathcal{C}$ will tell us the amount of material for each possible particle state. That is, each state is fully identified by a function $\rho_\mathcal{c}: \mathcal{S} \to \mathbbm{R}$ that returns the amount of material in the composite state $\mathcal{c}$ that is prepared in the particle state $\mathcal{s}$. For example, given a certain configuration of gas, we can tell the density of material that is at a specific point in space with a specific value of momentum.

@@

The ability to associate finite amounts of material with sets of states allows us to define finite regions of $\mathcal{S}$ and compare them. Consider a region of position and momentum in phase space. If we are able to spread a finite amount of material into a non-infinitesimal uniform distribution, then we know we have a finite region. And the density will give us an indication of how big the region is: if we double the region, the density will halve. In other words: because we are describing densities over $\mathcal{S}$, we are able to give a unique measure\footnote{Here measure is intended in the mathematical sense (a real valued function of a $\sigma$-algebra) which is distinct from other connotations in physics or probability theory.} $\mu$ for the size of sets of $\mathcal{S}$. In terms of said measure, positive linear functionals become integrals, $\Lambda_U (\mathcal{c}) = \int_U \rho_{\mathcal{c}} d \mu$.

@@

The first order of business is to guarantee that the density remains defined under an arbitrary change of state variables. This means the Jacobian of the transformation must exist. For the Jacobian to exist the transformation between state variables has to be differentiable. Therefore a manifold that guarantees densities to be well defined is one that guarantees that state variable transformations are differentiable: a differentiable manifold. $\mathcal{S}$ has a differential structure in the sense that the distributions and the volume element $d\mu$ are only defined on a set of state variables that are linked by differentiable transformations. In the same way that discontinuous changes of state variable can be detected because they tamper with physical distinguishability (i.e. the topology), non differentiable changes of state variable can be detected because they tamper with our ability to define state-variable-independent densities.

@@

Another consideration is that while the norm associated to $L^1(\mathcal{S})$ is $\int_{\mathcal{S}} |\rho| d\mu$, the expression $\int_{\mathcal{S}} \rho d\mu$ is perhaps more physically meaningful. For proper states, the two are the same and they represent the total amount of material. For state changes they differ. The second represents the amount of material added (or taken away if negative) by the state change. For example, if $\delta \rho$ is the change due to deterministic and reversible evolution, the amount of material does not change and therefore $\int_{\mathcal{S}} \delta \rho \, d\mu = 0$. The first norm would represent the total amount of material that is changing (i.e. being added and being removed), therefore $\int_{\mathcal{S}} | \delta \rho | \, d\mu = 0$ means no material is moving, no change is occurring.

Note that neither of those expressions matches the norm given by the inner product defined as $\langle f, g \rangle = \int_{\mathcal{S}} fg \, d\mu$. Moreover $\mathcal{C}$, since it excludes discontinuous functions, can never be a complete normed space: it will not include the limit for all Cauchy sequences; it cannot be a Banach or a Hilbert space. Such a construction, though, is still useful. Consider the following expressions:
\begin{align*}
\Lambda_{\mathcal{S}} (\mathcal{c}) &= \int_\mathcal{S} \rho_{\mathcal{c}} d \mu = \langle 1 , \rho_{\mathcal{c}} \rangle \\
\Lambda_U (\mathcal{c}) &= \int_U \rho_{\mathcal{c}} d \mu = \langle 1_U , \rho_{\mathcal{c}} \rangle \\
\rho_\mathcal{c}(q_0,p_0) &= \int_\mathcal{S} \delta_{q_0,p_0} \cdot \rho_{\mathcal{c}} d \mu = \langle \delta_{q_0,p_0} , \rho_{\mathcal{c}} \rangle \\
p_{total}(\mathcal{c}) &= \int_\mathcal{S} p \cdot \rho_{\mathcal{c}} d \mu = \langle p , \rho_{\mathcal{c}} \rangle \\
q_{avg}(\mathcal{c}) &= \frac{\int_\mathcal{S} q \cdot \rho_{\mathcal{c}} d \mu}{\int_\mathcal{S} \rho_{\mathcal{c}} d \mu} = \frac{\langle q , \rho_{\mathcal{c}} \rangle}{\langle 1 , \rho_{\mathcal{c}} \rangle}
\end{align*}
For each state, they represent respectively the total amount of material, the material within $U$, the density at $(q_0, p_0)$, the total momentum and the average of a state variable. As they are linear functionals of $\mathcal{C}$,  they can be expressed using their dual vector over $\mathcal{S}$. But these vectors ($1_U$, $p$, ...) are \emph{not} elements of $\mathcal{C}$. They do not represent states, they are not continuous and integrable, and there is no single general physical meaning for all of them. Therefore, as we mentioned before, while we can extend the function space for convenience, computation purposes or to study a limit case, we need to be mindful of the extension and carefully check that the mathematical results are physically meaningful case by case.

@@

Physically, this means that the state variable $k$, which is the classical analogue of the wave number, is expressed with the inverse of the unit used for $q$. If $q$ is in meters, $k$ is in inverse meters. Consider now the area $dq \wedge dk$ of an infinitesimal rectangular region. This quantity is dimensionless and therefore invariant. The total number of states will be proportional to it, as  doubling the range of $dq$ or $dk$ will give us double the number of possible states. We have $d\mu = \hbar dq \wedge dk$, where $\hbar$ is the proportionality constant that will depend on the unit chosen to count the possibilities of the pair $(q, k)$. That is, a unitary range of possibilities for $q$ and $k$ will give us $\hbar$ possibilities.\footnote{The actual value and physical dimensions of $\hbar$ are determined by the system of units, and should not be taken to describe some intrinsic physical property. Only dimensionless relationships to other physical constants, such as the fine structure constant $\alpha = e^2/\hbar c$, can possess that trait. That is why one can choose ``natural units" for which $\hbar=1$. In SI units the relationship $\hbar dk = dp = m du$ between kinetic and conjugate momentum, derived later, sets the relationship between units.} The cardinality for states in a finite (i.e. compact) region $U \subseteq \mathcal{S}$ will be given by $\mu(U) = \int_U \hbar dq \wedge dk$: we are finally able to write the volume of integration in terms of state variables.

We can generalize to the case where $\mathcal{Q}$ has more than one dimension. As we must be able to change one unit at a time, independently from the other, to each $q^i$ will correspond a $k_i$ that uses the inverse of the corresponding units. We have $\mathcal{S}\cong T^*\mathcal{Q}$ and $d\mu = \hbar^n dq^n \wedge dk_n$. But it's not just the volume that is preserved as we change state variables. Each pair $(q^i, k_i)$ is expressed by an independent unit with the cardinality of possibilities given by $\hbar dq^i \wedge dk_i$. We call such a pair a degree of freedom. As they are independent the total number of states is the product of the possibilities of each d.o.f.: $d\mu = \hbar^n dq^n \wedge dk_n = \prod \limits_{i=1}^n \hbar dq^i \wedge dk_i$. In other words: independent d.o.f. are orthogonal surfaces in $\mathcal{S}$. The possibilities of an arbitrary degree of freedom, then, will be $\sum \limits_{i=1}^n \hbar dq^i \wedge dk_i$, the sum of the projections over the $n$ orthogonal and independent d.o.f. defined by $(q^i, k_i)$.

The cardinality of possibilities for each d.o.f. (i.e. the wedge product within a d.o.f.) and the orthogonality of different d.o.f. (i.e. the scalar product across d.o.f.) must be the same regardless of the choice of state variables. We can express both requirements mathematically in a compact way. We first define conjugate momentum as $p_i=\hbar k_i$ and unified state variables as $\xi^a\equiv \{q^i, p_i\}$. Then we consider the canonical two-form $\omega =\Sigma \, dq^i \wedge dp_i$ given by the following components:
\begin{align*}
\omega_{ab} =  \left[
\begin{array}{cc}
0 & 1 \\
-1 & 0 \\
\end{array}
\right] \otimes I_n =
\left[
\begin{array}{cc}
0 & I_n \\
-I_n & 0 \\
\end{array}
\right] \\
\end{align*}
It returns the wedge product within a d.o.f. and the scalar product across. Requiring the invariance of this metric under state variable changes assures us the cardinality of states and possibilities is well defined.

@@

This gives us insight on the physical meaning of the geometrical structure $T^*\mathcal{Q}$. The canonical one-form $\theta=k dq$ represents the geometrical object we associate with each particle state. The canonical two-form $\omega$ is the metric that quantifies the possibilities described by two given state variables. Note, instead, that the relationship $\omega = - \hbar d\theta$, while mathematically true, has no clear physical meaning as we have not defined what the exterior derivative on a state actually represents.

We can capture the above discussion by stating that the state space $\mathcal{S}$ is the symplectic manifold $(T^*\mathcal{Q}, \omega)$, where $\mathcal{Q}$ is the manifold that defines the unit system. This is the only manifold that allows us to define state-variable-invariant densities. The metric allows us to measure the cardinality of possibilities on an arbitrary d.o.f.

@@

The discussion so far has been limited to the case where both the set of states and the evolution map never change throughout the evolution. This is too restrictive as there are very reasonable situations in which this does not hold.

>>>>

We can also conceptually combine two different processes into a single one.\footnote{That is, take the logical and between outcomes of different processes.} That is, having a way to measure quantity $x$ and a way to measure quantity $y$ gives us a way to measure the combination $(x,y)$. For ensembles, if we can measure the marginal distribution $\rho_x(x)$ and the marginal distribution $\rho_y(y)$, we know the joint distribution $\rho(x,y)$ has to be compatible with both. Note, though, that this does not provide a way to fully measure $\rho(x,y)$ as we know nothing about the correlation between the two variables.\footnote{The quantum case is similar: we measure marginal distributions and rule out states that are incompatible with them.} Formally, the states compatible with the outcomes of the combined process will be the intersections of the states compatible with each pair of outcomes of the original processes.

We can also coarsen a single process.\footnote{That is, take the logical or between outcomes of the same process.} That is, having a way to measure $(x,y)$ gives us a way to measure $x$ alone (or any $f(x,y)$). Formally, the outcomes of the coarsened process are given by performing the union of some outcomes of the original process.

@@

With our state space defined, deterministic and reversible evolution corresponds to a bijective map between initial and final states. But that's not enough. A physical process that can distinguish final states can be used, together with the evolution, to distinguish initial states. We prepare the initial state, let it evolve deterministically, measure the final state and use reversibility to convert the measure to one on the initial state. Therefore, not only the evolution is a bijection, but it also maps outcomes to outcomes. That is: the topology is mapped and preserved by the deterministic and reversible evolution because physical distinguishability must remain unchanged. The evolution is then a self-homeomorphism on the state space.

@@

The generality of this result explains why in physics one always assumes functions to be ``well behaved". As the result was derived from our notion of physical distinguishability, this is not a matter of practical convenience. Suppose we were able to prepare a force field that was zero everywhere in space except at a single point. This gives us a way to tag a specific point. But it also allows us to create an outcome compatible with only a single point. A finite precision measurement of the force provides us infinite precision of space, which we ruled out.\footnote{This assumes that we are able to position the probe perfectly, which we could do if we were able to manipulate forces at that precision.} This is what the math is telling us, that if we claim that outcomes can't distinguish isolated points (i.e. standard topology on $\mathbbm{R}^n$) then neither can maps. It is physical consistency that limits us to continuous functions.

@@

To identify and name states one uses a set of quantities, typically numbers. For example, the orbital of an electron in a hydrogen atom is identified by the quantum numbers $n$, $l$, $m$ and $s$. We call each of these quantities \emph{state variables}. We call a \emph{possibility} a possible value that can be taken by a state variable. We call a state variable \emph{discrete} or \emph{continuous} if the possibilities are integer or real numbers respectively. The topology used reflects the notion that integers can be measured perfectly (i.e. discrete topology) while real numbers can be measured up to a finite precision (i.e. standard topology on $\mathbbm{R}$) From now on, we are going to consider state spaces whose states can be identified, at least within a region, by a finite set of discrete and continuous state variables (i.e. the state space is locally isomorphic to $\bigcup\limits_{1 \leq i \leq n} \mathbbm{R}^{m_i}$).

We purposely use the term state variables instead of coordinates (even though that's what they are mathematically) as it would create confusion with space-time coordinates. We also avoid the term observable or measurable, as not all state variables may be directly physically tangible (e.g. conjugate momentum in a gauge theory). The only requirements for state variables is that they identify states. This means the possible values for state variables are physically distinguishable and are defined at equal time (since states are defined and mapped at a particular time).\footnote{This is the main reason that a quantity like velocity is not a suitable state variable, as it is  defined over an interval of time. Therefore velocity is always physically well defined but is not a state variable in general, while conjugate momentum is always a state variable but is not physically well defined by itself (it requires the vector potential to be specified as well). As we'll see later, when there exists a one to one map between velocity and conjugate momentum they can both be physically well defined state variables.}

@@

When combining multiple state variables, it is important to understand how they relate to each other. Consider the orbital of an electron in a hydrogen atom, which is identified by the quantum numbers $n$, $l$, $m$ and $s$. For each combination of $n$, $l$ and $m$, the spin $s$ can have two values. The choice of spin is independent from the rest. The choice of $l$, though, depends on the choice of $n$: for $n=1$ only $l=0$ is available; for $n=2$ we can choose $l=0$ or $l=1$. The choices are not independent. That is: two or more variables are independent if there always exists a state for any possible combination, if the total number of states is the product of the possibilities of each variable. We call \emph{state vector} a collection of independent state variables that fully identify a state.

@@

The second approach is to transport the state variables. The idea is to keep the connection to the initial state by labeling the future state by the original value of $q^i$, instead of by the future values. For example, the evolved state variable will not tell us the current value of pressure, but the one for the initial conditions. So we introduce a new set of variables for which $\hat{q}^i(\hat{\mathcal{s}})=q^i(\mathcal{s})$, which we can always do as the evolution is deterministic and reversible. The hyper-surfaces at constant $\hat{q}^i$ are the evolved hyper-surfaces at constant of $q^i$ (e.g. all the states that started with a particular value for pressure) which allows us to study how groups of states evolve in time. Given that the value of the transported state variables does not change during evolution, and that it is unique for each initial state, transported state variables also provide a way to label the trajectories themselves.

Evolved state variables are useful to write equations of motions, study how physical quantities change, form a physical picture of what happens. Transported state variables are useful to write invariants, study state space trajectories, form a geometric picture for the state space.\footnote{One should not confuse evolved/transported state variable with active/passive transformations or with Schroedinger/Heisenberg pictures. In those cases, the choice is between changing the state or the coordinates/observables. In our case, the state is always changing. The choice is between tracking the change with different values of the same state variable or different state variables that give the same value.}

@@

One reason that transported variables will be useful is that, during deterministic and reversible evolution, we need to make sure that the number of initial and final possibilities across independent state variables remains the same. This is more easily done using transported variable as the actual values do not change. In the discrete case, this is easy to show.

@@

The issue is that while deterministic evolution will map states to states, not all bijective state to state maps can be considered deterministic and reversible.\footnote{In fact, we have already seen that the map must at least be continuous.} This is a crucial point where this work is different from more usual treatments of dynamical systems. If one focuses only on the states themselves and counts them, it reaches the conclusion that a range of $1$m contains the same number of initial conditions of $1$Km: they have the same infinite number of elements. This conclusion has a number of problems: it does not match our physical intuition; it's not consistent with all types of topology; it does not explain why the laws of evolution are very often, not only continuous, but differentiable. In short: a bijection does not capture the entirety of the physical description.

If we work with a discrete topology then, by all means, the way to measure the size of a set is to count the elements in a set. But if we use the standard topology on $\mathbbm{R}$, isolated points have measure zero (i.e. we can't physically distinguish them): what defines the number of initial conditions is the range of the state variables. As we'll see, the geometrical structures in Hamiltonian/Lagrangian classical/quantum mechanics are there to allow us to properly count states and, more specifically, possibilities along independent state variables. If we think that counting states is equivalent to counting elements in the set, we miss the whole point of that geometrical structure.

@@

Given that scientific reductionism (i.e. the idea of reducing physical systems and interactions to the sum of their constituent parts in order to make them easier to study) is at the heart of fundamental physics, we now explore how to characterize a system in terms of its components. That is, we want to study the relationship between the state space of a composite system and the state space of its parts. In general, this is quite a complicated thing to do, which requires intimate knowledge of the system at hand. So we simplify our problem and study a material made of infinitesimal homogeneous parts. What we'll find is that under the additional assumption that the material is infinitesimally reducible (i.e. its state is equivalent to the states of its infinitesimal parts) and that each part undergoes deterministic and reversible evolution, the motion is suitably described by the standard framework of classical Hamiltonian mechanics.

\subsection{Homogeneous decomposable systems and vector spaces}
The notion that a system is decomposable means the states are equipped with a rule of composition that allows one to write $\mathcal{c}=\mathcal{c}_1+\mathcal{c}_2$: the composite system is the sum of its components. For example, the state of a ball is equal to the state of its top and bottom parts.

@@

\begin{justification}
We claim $\mathcal{C}$ is an additive monoid. There exists a law of composition $+ : \mathcal{C} \times \mathcal{C} \rightarrow \mathcal{C}$ that takes two states and returns one that is the physical composition of the two. The domain and codomain match because the material is homogeneous. The law is commutative $\mathcal{c}_1 +\mathcal{c}_2 = \mathcal{c}_2+\mathcal{c}_1$ and associative $(\mathcal{c}_1 + \mathcal{c}_2) + \mathcal{c}_3 = \mathcal{c}_1 + (\mathcal{c}_2 + \mathcal{c}_3)$, as it does not matter in what order we physically compose the components. There exists a unique zero element $\mathcal{c} + 0 = \mathcal{c}$ and it represents the physically empty state (i.e. no amount of material). $\mathcal{C}$ is an additive monoid by definition.

@@

We will call a \emph{particle} the smallest amount of a material to which we can assign an independent state. For example, a photon is a particle of light (it is the smallest amount of light we can describe), an infinitesimal amount of water is treated classically as a point particle (the smallest amount for a continuous fluid).\footnote{We use this definition as it allows us to retroactively talk on somewhat equal grounds about classical particles and quantum particles, and underscore how physically (and mathematically) they are very similar yet very different objects.}

We will call a \emph{classical material} one that is homogeneous and infinitesimally reducible. That is, we can keep decomposing the system indefinitely into smaller and smaller parts, each with a well defined state. In this case, the particles are the infinitesimal parts given by the limit of this process of recursive reduction. Giving the state of the whole system is equivalent to giving the states of all particles. Given the state space $\mathcal{S}$ for the infinitesimal parts, which we assume consistently with \ref{continuous_state_space} to be a manifold, a composite state $\mathcal{c} \in \mathcal{C}$ will tell us the amount of material for each possible particle state. That is, each state is fully identified by a function $\rho_\mathcal{c}: \mathcal{S} \to \mathbbm{R}$ that returns the amount of material in the composite state $\mathcal{c}$ that is prepared in the particle state $\mathcal{s}$. For example, given a certain configuration of gas, we can tell the density of material that is at a specific point in space with a specific value of momentum.

@@

The ability to associate finite amounts of material with sets of states allows us to define finite regions of $\mathcal{S}$ and compare them. Consider a region of position and momentum in phase space. If we are able to spread a finite amount of material into a non-infinitesimal uniform distribution, then we know we have a finite region. And the density will give us an indication of how big the region is: if we double the region, the density will halve. In other words: because we are describing densities over $\mathcal{S}$, we are able to give a unique measure\footnote{Here measure is intended in the mathematical sense (a real valued function of a $\sigma$-algebra) which is distinct from other connotations in physics.} $\mu$ that allows to count the number of states in each set of $\mathcal{S}$. In terms of said measure, positive linear functionals become integrals, $\Lambda_U (\mathcal{c}) = \int_U \rho_{\mathcal{c}} d \mu$.

@@

The first order of business is to guarantee that the density remains defined under an arbitrary change of state variables. This means the Jacobian of the transformation must exist. For the Jacobian to exist the transformation between state variables has to be differentiable. Therefore a manifold that guarantees densities to be well defined is one that guarantees that state variable transformations are differentiable: a differentiable manifold. $\mathcal{S}$ has a differential structure in the sense that the distributions and the volume element $d\mu$ are only defined on a set of state variables that are linked by differentiable transformations. In the same way that discontinuous changes of state variable can be detected because they tamper with physical distinguishability (i.e. the topology), non differentiable changes of state variable can be detected because they tamper with our ability to define state-variable-independent densities and count states.

@@

Another consideration is that while the norm associated to $L^1(\mathcal{S})$ is $\int_{\mathcal{S}} |\rho| d\mu$, the expression $\int_{\mathcal{S}} \rho d\mu$ may be more physically meaningful in some cases. For proper states, the two are the same and they represent the total amount of material. For state changes they differ. The second represents the amount of material added (or taken away if negative) by the state change. For example, if $\delta \rho$ is the change due to deterministic and reversible evolution, the amount of material does not change and therefore $\int_{\mathcal{S}} \delta \rho \, d\mu = 0$. The first norm would represent the total amount of material that is changing (i.e. being added and being removed), therefore $\int_{\mathcal{S}} | \delta \rho | \, d\mu = 0$ means no material is moving, no change is occurring.

Note that neither of those expressions matches the norm given by the inner product defined as $\langle f, g \rangle = \int_{\mathcal{S}} fg \, d\mu$. Moreover $\mathcal{C}$, since it excludes discontinuous functions, can never be a complete normed space: it will not include the limit for all Cauchy sequences; it cannot be a Banach or a Hilbert space. Such a construction, though, is still useful. Consider the following expressions:
\begin{align*}
\Lambda_{\mathcal{S}} (\mathcal{c}) &= \int_\mathcal{S} \rho_{\mathcal{c}} d \mu = \langle 1 , \rho_{\mathcal{c}} \rangle \\
\Lambda_U (\mathcal{c}) &= \int_U \rho_{\mathcal{c}} d \mu = \langle 1_U , \rho_{\mathcal{c}} \rangle \\
\rho_\mathcal{c}(q_0,p_0) &= \int_\mathcal{S} \delta_{q_0,p_0} \cdot \rho_{\mathcal{c}} d \mu = \langle \delta_{q_0,p_0} , \rho_{\mathcal{c}} \rangle \\
p_{total}(\mathcal{c}) &= \int_\mathcal{S} p \cdot \rho_{\mathcal{c}} d \mu = \langle p , \rho_{\mathcal{c}} \rangle \\
q_{avg}(\mathcal{c}) &= \frac{\int_\mathcal{S} q \cdot \rho_{\mathcal{c}} d \mu}{\int_\mathcal{S} \rho_{\mathcal{c}} d \mu} = \frac{\langle q , \rho_{\mathcal{c}} \rangle}{\langle 1 , \rho_{\mathcal{c}} \rangle}
\end{align*}
For each state, they represent respectively the total amount of material, the material within $U$, the density at $(q_0, p_0)$, the total momentum and the average of a state variable. As they are linear functionals of $\mathcal{C}$,  they can be expressed using their dual vector. But these vectors ($1_U$, $p$, ...) are \emph{not} elements of $\mathcal{C}$. They do not represent states, they are not continuous and integrable, and there is no single general physical meaning for all of them. Therefore, as we mentioned before, while we can extend the function space for convenience, computation purposes or to study a limit case, we need to be mindful of the extension and carefully check that the mathematical results are physically meaningful case by case.

@@

Physically, this means that the state variable $k$, which is the classical analogue of the wave number, is expressed with the inverse of the unit used for $q$. If $q$ is in meters, $k$ is in inverse meters. Consider now the area $dq \wedge dk$ of an infinitesimal rectangular region. This quantity is dimensionless and therefore invariant. The total number of states will be proportional to it, as  doubling the range of $dq$ or $dk$ will give us double the number of possible states. We have $d\mu = \hbar dq \wedge dk$, where $\hbar$ is the proportionality constant that will depend on the unit chosen to count the possibilities of the pair $(q, k)$. That is, a unitary range of possibilities for $q$ and $k$ will give us $\hbar$ possibilities.\footnote{The actual value and physical dimensions of $\hbar$ are determined by the system of units, and should not be taken to describe some intrinsic physical property. Only dimensionless relationships to other physical constants, such as the fine structure constant $\alpha = e^2/\hbar c$, can possess that trait. That is why one can choose ``natural units" for which $\hbar=1$. In SI units the relationship $\hbar dk = dp = m du$ between kinetic and conjugate momentum, derived later, sets the relationship between units.} The state count in a finite (i.e. compact) region $U \subseteq \mathcal{S}$ will be given by $\mu(U) = \int_U \hbar dq \wedge dk$: we are finally able to write the volume of integration in terms of state variables.

We can generalize to the case where $\mathcal{Q}$ has more than one dimension. As we must be able to change one unit at a time, independently from the other, to each $q^i$ will correspond a $k_i$ that uses the inverse of the corresponding units. We have $\mathcal{S}\cong T^*\mathcal{Q}$ and $d\mu = \hbar^n dq^n \wedge dk_n$. But it's not just the volume, the state count, that is preserved as we change state variables. Each pair $(q^i, k_i)$ is expressed by an independent unit with the count of possibilities given by $\hbar dq^i \wedge dk_i$. We call such a pair a degree of freedom. As they are independent the total state count is the product of the possibilities of each d.o.f.: $d\mu = \hbar^n dq^n \wedge dk_n = \prod \limits_{i=1}^n \hbar dq^i \wedge dk_i$. In other words: independent d.o.f. are orthogonal surfaces in $\mathcal{S}$. The possibility count of an arbitrary degree of freedom, then, will be $\sum \limits_{i=1}^n \hbar dq^i \wedge dk_i$, the sum of the projections over the $n$ orthogonal and independent d.o.f. defined by $(q^i, k_i)$.

The possibility count for each d.o.f. (i.e. the wedge product within a d.o.f.) and the orthogonality of different d.o.f. (i.e. the scalar product across d.o.f.) must be the same regardless of the choice of state variables. We can express both requirements mathematically in a compact way. We first define conjugate momentum as $p_i=\hbar k_i$ and unified state variables as $\xi^a\equiv \{q^i, p_i\}$. Then we consider the canonical symplectic two-form $\omega =\Sigma \, dq^i \wedge dp_i$ given by the following components:
\begin{align*}
\omega_{ab} =  \left[
\begin{array}{cc}
0 & 1 \\
-1 & 0 \\
\end{array}
\right] \otimes I_n =
\left[
\begin{array}{cc}
0 & I_n \\
-I_n & 0 \\
\end{array}
\right] \\
\end{align*}
It returns the wedge product within a d.o.f. and the scalar product across. Requiring the invariance of this metric under state variable changes assures us the count of states and possibilities is well defined.

@@

This gives us insight on the physical meaning of the geometrical structure $T^*\mathcal{Q}$. The canonical one-form $\theta=k dq$ represents the geometrical object we associate with each particle state. The symplectic form $\omega$ is the bilinear that quantifies the possibilities described by two given state variables. Note, instead, that the relationship $\omega = - \hbar d\theta$, while mathematically true, has no clear physical meaning as we have not defined what the exterior derivative on a state actually represents.

We can capture the above discussion by stating that the state space $\mathcal{S}$ is the symplectic manifold $(T^*\mathcal{Q}, \omega)$, where $\mathcal{Q}$ is the manifold that defines the unit system. This is the only manifold that allows us to define state-variable-invariant densities. The symplectic form allows us to count possibilities on an arbitrary d.o.f.

@@

The discussion so far has been limited to the case where both the state variables and the evolution map never change throughout the evolution. This is too restrictive as there are very reasonable situations in which this does not hold.
